---
title: "TFM - Estudi demogràfic"
author: "Salvador Sanchis Beneseit"
date: '`r format(Sys.Date(),"%B %e, %Y")`'
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, include=FALSE}
library(knitr)
```

------------------------------------------------------------------------

<br>

## 1. Data loading and preprocessing

<br>

We load the data:

```{r}
clinic <- read.csv("clinic.csv", sep=",")
```

<br>

We filter our variables of interest, only from ms patients, into
`clinic_data` :

```{r}
clinic_ms <- subset(clinic, controls_ms == 1)
columns_to_keep <- c("id", "age", "sex", "dd", "mstype","edss")
clinic_data <- clinic_ms[, columns_to_keep]
```

<br>

```{r}
summary(clinic_data)
```

<br>

We check and correct data formats:

```{r}
sapply(clinic_data, class)
```

<br>

```{r}
clinic_data$sex[clinic_data$sex == 0] <- "F"
clinic_data$sex[clinic_data$sex == 1] <- "M"
clinic_data$sex <- as.factor(clinic_data$sex)
levels(clinic_data$sex)

```

<br>

```{r}
clinic_data$mstype[clinic_data$mstype == 0] <- "RRMS"
clinic_data$mstype[clinic_data$mstype == 1] <- "SPMS"
clinic_data$mstype[clinic_data$mstype == 2] <- "PPMS"
clinic_data$mstype <- as.factor(clinic_data$mstype)
levels(clinic_data$mstype)
```

<br>

```{r}
sapply(clinic_data, class)
```

<br>

## 2. Visual inspection of data

<br>

```{r}
boxplot(clinic_data$age, main="Age Distribution MS patients")
```

<br>

```{r}
library(ggplot2)
ageplot <- ggplot(clinic_data, aes(age))
ageplot +geom_histogram(binwidth = 1, colour = 'green', size=0.2) + ggtitle("Age distribution of MS patients")
```

<br>

```{r}
ddplot <- ggplot(clinic_data, aes(dd))
ddplot +geom_histogram(binwidth = 1, colour = 'green', size=0.2) + ggtitle("Disease duration of MS patients")
```

<br>

```{r}
edssplot <- ggplot(clinic_data, aes(edss))
edssplot +geom_histogram(binwidth = 0.5, colour = 'green', size=0.2) + ggtitle("Distribution of EDSS scores")
```

<br>

```{r}
gender_count <- table(clinic_data$sex)
barplot(gender_count, main="Gender of MS patients")
```

<br>

```{r}
ms_type_count <- table(clinic_data$mstype)
barplot(ms_type_count, main="MS phenotype count")
```

<br>

```{r}
ggplot(clinic_data, aes(x = mstype, fill = sex)) +
  geom_bar(position = "dodge", stat = "count") +
  labs(title = "Counts of ms type grouped by Gender", x = "mstype", y = "Count") +
  scale_fill_manual(values = c("darkolivegreen", "darkmagenta"))  
```

\
<br>

In this plot we see no pattern to the correlation between disease
duration and EDSS score:

```{r}
plot(clinic_data$dd, clinic_data$edss, 
     main = "Scatter Plot of Disease Duration vs. EDSS",
     xlab = "Disease Duration (years)",
     ylab = "EDSS")
```

<br>

However, introducing phenotype information, we see that high edss scores
correspond mainly to PPMS and SPMS types, with lower and higher disease
duration respectively:

```{r}
ggplot(clinic_data, aes(x = dd, y = edss, color = mstype)) +
  geom_point() +
  labs(title = "Scatter Plot of Disease Duration vs. EDSS and MS phenotype",
       x = "Disease Duration (years)",
       y = "EDSS")
```

<br>

In the following cells, we explore relations between brain volumes, EDSS
and type of disease. Following our state of the art preliminary work, we
are curious about the role of atrophy of the thalamus in MS.

<br>

```{r}

# We subset data from left and right thalamus and we eliminate outliers
volumes_controls <- read.csv('volum_nodes_controls.csv')
rownames(volumes_controls) <- volumes_controls$X0
volumes_controls <- volumes_controls[, -1]

volumes_patients <- read.csv('volum_nodes_patients.csv')
rownames(volumes_patients) <- volumes_patients$X0
volumes_patients <- volumes_patients[, -1]

all_volumes <- rbind(volumes_patients, volumes_controls)

clinic_data['avg_volumes'] <- rowMeans(volumes_patients)

thalamus_all <- all_volumes[, c('X32', 'X39')]

```

```{r}
boxplot(thalamus_all)
```

```{r}
outlier <- which.min(thalamus_all$X39)
thalamus_corrected <- thalamus_all[-outlier, ]
boxplot(thalamus_corrected)
```

```{r}
clinic_data_th <- clinic_data[-78, ]
clinic_data_th['left_thalamus'] <- thalamus_corrected$X32[1:146]
clinic_data_th['right_thalamus'] <- thalamus_corrected$X39[1:146]
clinic_data_th['thal_avg'] <- (clinic_data_th$right_thalamus + clinic_data_th$left_thalamus) / 2

```

<br>

In this plot we relate average brain volumes (all 76 nodes) with EDSS
and disease phenotype:

```{r}
library(ggplot2)
ggplot(clinic_data_th, aes(x = avg_volumes, y = edss, color = mstype)) +
  geom_point() +
  labs(title = "Scatter Plot of Average Brain Volumes vs. EDSS and MS phenotype",
       x = "Avg_volumes",
       y = "EDSS")

```

<br>

Next, we do similar comparison, but this time using only thalamic
volumes (we switch the axes of the visualization):

```{r}
ggplot(clinic_data_th, aes(x = edss, y = thal_avg, color = mstype)) +
  geom_point() +
  labs(title = "Scatter Plot",
       x = "EDSS",
       y = "Thal")
```

<br>

As the 2 previous visualizations are inconclusive, we show a comparison
of boxplots for the different levels of EDSS in relation to thalamic
volume:

```{r}
ggplot(aes(y=thal_avg, x=as.factor(edss)), data=clinic_data_th) + geom_boxplot()
```

<br>

Since we perceive a slightly descending trend in the graph, we decide to
do an ANOVA test to see if there is a significant relation between
thalamic volumes and EDSS. The results show that this is indeed the case
(p=0.0025), so we will include thalamic volumes in our classification
task.

```{r}
lm_thal <- lm(thal_avg~as.factor(edss), data = clinic_data_th)
table_anova <- anova(lm_thal)
table_anova
```

<br>

```{r}
```

```{r}

ggplot(clinic_data_th, aes(x = thal_avg, y = as.numeric(edss), color = mstype)) +
  geom_point() + geom_smooth(method = 'lm', se=FALSE)



```

<br>

## 3. Statistical measures

<br>

In the following cells, we describe groups and study their statistical
differences (patients vs. controls, and disease phenotype groups) in
terms of age, disease duration, EDSS and gender. All results are
summarized on a table in the last section of this notebook.

<br>

```{r}
#we define a function to calculate median and iqr
calculate_summary_stats <- function(data, condition) {
  vector <- condition
  result <- list(
    median = median(vector),
    q1 = quantile(vector, 0.25),
    q3 = quantile(vector, 0.75)
  )
  formatted_result <- sprintf("%.2f (%.2f - %.2f)", result$median, result$q1, result$q3)
  
  return(formatted_result)
}

```

<br>

### Age

```{r}
hc_age <- calculate_summary_stats(clinic, clinic$age[clinic$controls_ms == 0])
rrms_age <- calculate_summary_stats(clinic_data, clinic_data$age[clinic_data$mstype == "RRMS"])
spms_age <- calculate_summary_stats(clinic_data, clinic_data$age[clinic_data$mstype == "SPMS"])
ppms_age <- calculate_summary_stats(clinic_data, clinic_data$age[clinic_data$mstype == "PPMS"])

hc_age
rrms_age
spms_age
ppms_age
```

<br>

```{r}
hc_age_data <- clinic$age[clinic$controls_ms == 0]
ms_age_data <- clinic$age[clinic$controls_ms == 1]

wilcoxon_age <- wilcox.test(hc_age_data, ms_age_data)
formatted_p_value_w <- ifelse(wilcoxon_age$p.value < 0.001, "p < 0.001", sprintf("p = %.3f", wilcoxon_age$p.value))
wilcoxon_age_p <- c(wilcoxon_age, Formatted_P_Value = formatted_p_value_w)
wilcoxon_age_p
```

```{r}
rrms_age_data <- clinic_data$age[clinic_data$mstype == "RRMS"]
spms_age_data <- clinic_data$age[clinic_data$mstype == "SPMS"]
ppms_age_data <- clinic_data$age[clinic_data$mstype == "PPMS"]

# Kruskal-Wallis test
kruskal_age <- kruskal.test(list(rrms_age_data, spms_age_data, ppms_age_data))
formatted_p_value <- ifelse(kruskal_age$p.value < 0.001, "p < 0.001", sprintf("p = %.3f", kruskal_age$p.value))
kruskal_age_p <- c(kruskal_age, Formatted_P_Value = formatted_p_value)
kruskal_age_p
```

```{r}
library(dunn.test)

# Perform Dunn's test for pairwise comparisons
dunn_age <- dunn.test(list(rrms_age_data, spms_age_data, ppms_age_data), method = "bonferroni")
# The results show difference between groups, except for comparison between spms and ppms
dunn_age
```

<br>

### Disease duration

<br>

```{r}
rrms_dd <- calculate_summary_stats(clinic_data, clinic_data$dd[clinic_data$mstype == "RRMS"])
spms_dd <- calculate_summary_stats(clinic_data, clinic_data$dd[clinic_data$mstype == "SPMS"])
ppms_dd <- calculate_summary_stats(clinic_data, clinic_data$dd[clinic_data$mstype == "PPMS"])

rrms_dd
spms_dd
ppms_dd
```

```{r}
rrms_dd_data <- clinic_data$dd[clinic_data$mstype == "RRMS"]
spms_dd_data <- clinic_data$dd[clinic_data$mstype == "SPMS"]
ppms_dd_data <- clinic_data$dd[clinic_data$mstype == "PPMS"]

# Kruskal-Wallis test
kruskal_dd <- kruskal.test(list(rrms_dd_data, spms_dd_data, ppms_dd_data))
formatted_p_value_dd <- ifelse(kruskal_dd$p.value < 0.001, "p < 0.001", sprintf("p = %.3f", kruskal_dd$p.value))
kruskal_dd_p <- c(kruskal_dd, Formatted_P_Value = formatted_p_value_dd)
kruskal_dd_p

```

```{r}
# Perform Dunn's test for pairwise comparisons
dunn_dd <- dunn.test(list(rrms_dd_data, spms_dd_data, ppms_dd_data), method = "bonferroni")
# The results show difference between groups, except for comparison between rrms and ppms
dunn_dd
```

<br>

### EDSS

<br>

```{r}
rrms_edss <- calculate_summary_stats(clinic_data, clinic_data$edss[clinic_data$mstype == "RRMS"])

spms_edss <- calculate_summary_stats(clinic_data, clinic_data$edss[clinic_data$mstype == "SPMS"])

ppms_edss <- calculate_summary_stats(clinic_data, clinic_data$edss[clinic_data$mstype == "PPMS"])

spms_edss
rrms_edss
ppms_edss
```

```{r}
rrms_edss_data <- clinic_data$edss[clinic_data$mstype == "RRMS"]
spms_edss_data <- clinic_data$edss[clinic_data$mstype == "SPMS"]
ppms_edss_data <- clinic_data$edss[clinic_data$mstype == "PPMS"]

# Kruskal-Wallis test
kruskal_edss <- kruskal.test(list(rrms_edss_data, spms_edss_data, ppms_edss_data))
formatted_p_value_edss <- ifelse(kruskal_edss$p.value < 0.001, "p < 0.001", sprintf("p = %.3f", kruskal_edss$p.value))
kruskal_edss_p <- c(kruskal_edss, Formatted_P_Value = formatted_p_value_edss)
kruskal_edss_p
```

```{r}
# Perform Dunn's test for pairwise comparisons
dunn_edss <- dunn.test(list(rrms_edss_data, spms_edss_data, ppms_edss_data), method = "bonferroni")
# The results show difference between groups, except for comparison between spms and ppms
dunn_edss
```

<br>

### Gender

<br>

```{r}
calculate_gender_stats <- function(data, group_column, gender_column, group_value) {
  # Subset data for the specified group
  group_data <- data[data[[group_column]] == group_value, ]
  
  # Count of females
  female_count <- sum(group_data[[gender_column]] == 0 | group_data[[gender_column]] == "F")
  
  total_count <- nrow(group_data)
  # Percentage of females
  female_percentage <- (female_count / total_count) * 100
  
  result <- list(
    Female_Count = female_count,
    Female_Percentage = female_percentage
  )
  formatted_result <- sprintf("%d (%.f%%)", result$Female_Count, result$Female_Percentage)
  
  return(formatted_result)
}
```

<br>

```{r}
female_hv <- calculate_gender_stats(clinic, "controls_ms", "sex", 0)
female_rrms <- calculate_gender_stats(clinic_data, "mstype", "sex", "RRMS")
female_spms <- calculate_gender_stats(clinic_data, "mstype", "sex", "SPMS")
female_ppms <- calculate_gender_stats(clinic_data, "mstype", "sex", "PPMS")
female_hv
female_rrms
female_spms
female_ppms
```

<br>

```{r}
gender_table_controls <- table(clinic$controls_ms == 0, clinic$sex)
gender_table_controls
chi_sq_ms <- chisq.test(gender_table_controls)
chi_sq_ms
```

<br>

```{r}
gender_table_phenotypes <- table(clinic_data$mstype, clinic_data$sex)
gender_table_phenotypes
chi_sq_types <- chisq.test(gender_table_phenotypes)
chi_sq_types
```

<br>

## Summary

<br>

```{r}
summary_table <- data.frame(HCs=c(hc_age, "-", "-", female_hv),
                            RRMS=c(rrms_age, rrms_dd, rrms_edss, female_rrms),
                            SPMS=c(spms_age, spms_dd, spms_edss, female_spms),
                            PPMS=c(ppms_age, ppms_dd, ppms_edss, female_ppms),
                            p_value_HC_vs_MS=c(wilcoxon_age_p$Formatted_P_Value, "-", "-", round(chi_sq_ms$p.value, 3)),
                            p_value_MS_types=c(kruskal_age_p$Formatted_P_Value, kruskal_dd_p$Formatted_P_Value, kruskal_edss_p$Formatted_P_Value, round(chi_sq_types$p.value, 3)),
                            row.names = c('Age(IQR)', 'Disease duration', 'EDSS', 'Female(%)'))

summary_table
```

<br>\
